---
title: "xG Timeline Plots"
output: github_document
---

## xG Accumulative Charts 

Expected Goals (xG) is a popularised metric for calculating the probabilty that a particular shot will be scored - explainer https://www.bbc.co.uk/sport/football/40699431. xG is a more reliable metric when assessed over seasons worth of data and single-game analysis using xG has huge weaknesses. However, single-game accumulative analysis has become popular, mainly because it shows a narrative of the match over time. 

Using Statsbomb free women's football data I will run through the code to create the xG Accumulative plots. 

## The Data & Setup 

I will be using the recently released dataset from [Statsbomb](https://twitter.com/statsbomb), which is free to use and [easy to access](https://www.github.com). The dataset consists of 11 games at the time of writing, which is equivilent to over 25,000 events. 

#### Data Load
```{r}
events <- readRDS("SB_events_DB.RDS") ## an events dataframe loaded from local storage - see my previous tutorial 
```

#### Package Loading
```{r}
require(dplyr); require(ggplot2)
```

#### Filtering to one match and two teams 
```{r}
## pick a random match from the Statsbomb dataset 
randomMatch <- unique(events$match_id)[runif(1, 1, length(unique(events$match_id)) )]

## select just the shots from that match 
shotData <- events %>% filter(match_id == randomMatch & shot.statsbomb_xg > 0)

## create a more precise metric for 'time' which combines minutes and seconds 
shotData$Timer <- shotData$minute + (shotData$second / 60)

## drop all unneeed data to make it easier to work with the data 
shotData <- shotData[c(19,54,61,147)]

## Create a dataframe just for the Team 1 and Team 2
T1_shotData <- shotData %>% filter(team.name == unique(shotData$team.name)[1])
T2_shotData <- shotData %>% filter(team.name == unique(shotData$team.name)[2])
```

#### Create the accumulative values for xG for each team  

We calculate the accumulative xG total using the cumsum() function which comes pre-installed with R install. This gives us back just the data we need.

```{r}
## Team 1
# Calculate the accumulative xG total and add it as a column to the dataframe 
T1_shotData$xg_total <- cumsum(T1_shotData$shot.statsbomb_xg)
# create the starting and ending rows of data to ensure that all teams start on 0 and end at 96 minutes 
Top2add <- data.frame(team.name = unique(shotData$team.name)[1], shot.statsbomb_xg = 0, Timer = 0, shot.outcome.name = "-", xg_total = 0, stringsAsFactors = F)
Bottom2add <- data.frame(team.name = unique(shotData$team.name)[1], shot.statsbomb_xg = 0, Timer = 96,  shot.outcome.name = "-", xg_total = max(T1_shotData$xg_total), stringsAsFactors = F)

## add these rows to the data with bind_rows()
T1_shotData <- bind_rows(T1_shotData, Top2add)
T1_shotData <- bind_rows(T1_shotData, Bottom2add)


## Team 2
# Calculate the accumulative xG total and add it as a column to the dataframe 
T2_shotData$xg_total <- cumsum(T2_shotData$shot.statsbomb_xg)
# create the starting and ending rows of data to ensure that all teams start on 0 and end at 96 minutes 
Top2add <- data.frame(team.name = unique(shotData$team.name)[2], shot.statsbomb_xg = 0, Timer = 0,  shot.outcome.name = "-", xg_total = 0, stringsAsFactors = F)
Bottom2add <- data.frame(team.name = unique(shotData$team.name)[2], shot.statsbomb_xg = 0, Timer = 96,  shot.outcome.name = "-", xg_total = max(T2_shotData$xg_total), stringsAsFactors = F)

## add these rows to the data with bind_rows()
T2_shotData <- bind_rows(T2_shotData, Top2add)
T2_shotData <- bind_rows(T2_shotData, Bottom2add)
```


## Plot the Data 

Now we have all the data to plot lets build our plots, as always I build it layer by layer to help learning. We will use the trusted ggplot2() package. 

We will use the geom_step() function that is designed specifically for the type of chart we want and takes all the hard work out of it!

#### Basic Plot 
```{r}
p <- ggplot()
p
```

#### Add Team 1 Data 
```{r}
p <- p + geom_step(data=T1_shotData, mapping=aes(x=Timer, y=xg_total), colour = "#E24F55", size = 1.5, alpha = 0.8)
p
```

#### Add Team 1 Data 
```{r}
p <- p + geom_step(data=T2_shotData, mapping=aes(x=Timer, y=xg_total), colour = "#2B6DD2", size = 1.5, alpha = 0.8)
p
```


#### Add the Goals 

```{r}
### prepare the data 
T1_goals <- T1_shotData %>% filter(shot.outcome.name == "Goal")
T2_goals <- T2_shotData %>% filter(shot.outcome.name == "Goal")

## add the points on for the goals 
p <- p + 
geom_point(data = T1_goals, aes(x=Timer, y=xg_total), colour = "black", size = 4, alpha = 1, fill = "#E24F55", shape = 21) +
geom_point(data = T2_goals, aes(x=Timer, y=xg_total), colour = "black", size = 4, alpha = 1, fill = "#2B6DD2", shape = 21)

p
```

#### Labels, Theme & Legends  
```{r}
## find the axis length for the plot
yMax <- ggplot_build(p)$layout$panel_ranges[[1]]$y.range

## create the annotation titles 
T1_label <- paste0(unique(shotData$team.name)[1], ": ", nrow(T1_goals), " Actual: ", round(max(T1_shotData$xg_total), 2), "xG Diff: ", nrow(T1_goals) - round(max(T1_shotData$xg_total), 2))

T2_label <- paste0(unique(shotData$team.name)[2], ": ", nrow(T2_goals), " Actual: ", round(max(T2_shotData$xg_total), 2), "xG Diff: ", nrow(T2_goals) - round(max(T2_shotData$xg_total), 2))

## add the legend 
p + 
annotate("text", x = 0, y = (yMax[2] - (yMax[2]*0.05)), label = T1_label, hjust = 0, colour = "#E24F55") + 
annotate("text", x = 0, y = (yMax[2] - (yMax[2]*0.1)), label = T2_label, hjust = 0, colour = "#2B6DD2") + labs(x = "Minutes", y = "Accumulative xG") + theme_minimal()

```

There are many ways to style these plots but the above code will get you to the final product for you to tweak further.. enjoy. 